---
title: "DataExploration"
format: html
editor: visual
---

Setting Up

```{r}
rm(list=ls())
library(tidyverse)
library(here)
library(ggfortify)
library(multcomp)
```

Importing data

```{r}
KelpForest <- read.csv(here("Data", "KelpForestBiomassSurvey.csv"))
glimpse(KelpForest)
```





Date needs to be changed to a date data type. 
```{r}
KelpForest$DATE <- ymd(KelpForest$DATE)
```


Site needs to be converted to factor data since it will be used as a differentiation variable. 
```{r}
KelpForest$SITE <- as.factor(KelpForest$SITE)
levels(KelpForest$SITE)#looks good
```

SCIENTIFIC_NAME needs to be converted to factor data since I am going to use it to identify species richness 
```{r}
KelpForest$SCIENTIFIC_NAME <- as.factor(KelpForest$SCIENTIFIC_NAME)
levels(KelpForest$SCIENTIFIC_NAME)
```

I did not level the other categorizing columns for the species since I won't be using them as of now to do any data analysis. 

NA values need to be changed to na instead of -99999999

Using PERCENT_COVER < -1 also gets rid of any 

```{r}
KelpForest <- KelpForest |> mutate(PERCENT_COVER = 
                                     ifelse(PERCENT_COVER < -1, NA, PERCENT_COVER))

KelpForest <- KelpForest |> mutate(DENSITY = 
                                     ifelse(DENSITY < -1, NA, DENSITY))

KelpForest <- KelpForest |> mutate(WM_GM2 = 
                                     ifelse(WM_GM2 < -1, NA, WM_GM2))

KelpForest <- KelpForest |> mutate(DRY_GM2 = 
                                     ifelse(DRY_GM2 < -1, NA, DRY_GM2))

KelpForest <- KelpForest |> mutate(SFDM = 
                                     ifelse(SFDM < -1, NA, SFDM))

KelpForest <- KelpForest |> mutate(AFDM = 
                                     ifelse(AFDM < -1, NA, AFDM))

KelpForest <- KelpForest |> mutate(AFDM = 
                                     ifelse(AFDM < -1, NA, AFDM))

KelpForest <- KelpForest |> mutate(COMMON_NAME = 
                                     ifelse(COMMON_NAME == "-99999", NA, COMMON_NAME))

KelpForest <- KelpForest |> mutate(TAXON_KINGDOM = 
                                     ifelse(TAXON_KINGDOM == "-99999", NA, TAXON_KINGDOM))

KelpForest <- KelpForest |> mutate(TAXON_PHYLUM = 
                                     ifelse(TAXON_PHYLUM == "-99999", NA, TAXON_PHYLUM))

KelpForest <- KelpForest |> mutate(TAXON_CLASS = 
                                     ifelse(TAXON_CLASS == "-99999", NA, TAXON_CLASS))

KelpForest <- KelpForest |> mutate(TAXON_ORDER = 
                                     ifelse(TAXON_ORDER == "-99999", NA, TAXON_ORDER))

KelpForest <- KelpForest |> mutate(TAXON_FAMILY = 
                                     ifelse(TAXON_FAMILY == "-99999", NA, TAXON_FAMILY))

KelpForest <- KelpForest |> mutate(TAXON_GENUS = 
                                     ifelse(TAXON_GENUS == "-99999", NA, TAXON_GENUS))
```

I found the ifelse function while looking online for how to change just certain values since it wasn't working when I was trying to filter values for the double point data type columns. I just kept using it with my character data columns to keep my code consistent even though I think using filter piped into mutate would work there.

Doing summary stats and checking for outliers.

I am anticipating that actually removing any outlines won't be feasible since there will be wide ranges between species presence I expect. 

Making a scatterplot of PERCENT_COVER

```{r}
ggplot(KelpForest, aes(PERCENT_COVER, DATE))+
  geom_point()
```
using filter to look at the outliers and see if they should be removed or not
```{r}
KelpForest |> filter(PERCENT_COVER > 75)
```
I did some basic research into these two species and their percent cover seems pretty expected and nothing out of the ordinary so I will be leaving them in the data. 


DENSITY
Making a scatterplot
```{r}
ggplot(KelpForest, aes(DENSITY, DATE))+
  geom_point()
```
using filter to look at the higher density points. 
```{r}
KelpForest |> filter(DENSITY > 90)
```

My online research didn't give any results that would make me think either of these organisms couldn't occur in such densities so I am going to leave them in the data. 

DRY_GM2
Making a scattorplot
```{r}
ggplot(KelpForest, aes(DRY_GM2, DATE))+
  geom_point()
```
using filter to look at the one outlier
```{r}
KelpForest |> filter(DRY_GM2 > 30000)
```

They can grow to about 1.5-2 pounds each. Not really sure enough about their biology to take them out so I will leave them in for now and consult with you in class. 


Make some box plots and summary stats for the tests I am going to be running. 
```{r}
ggplot(KelpForest, aes(SITE, DRY_GM2))+
  geom_boxplot()+
  facet_wrap(~ YEAR)
```
The plots really don't show much since there is such a larger variability in the data so it is hard to tell. Going to just use summary statistics. 

```{r}
KelpForest |> group_by(YEAR) |>
  summarise(mean_drymass = mean(DRY_GM2, na.rm = TRUE))
```
There does not seem to be any specific trend over time for mean dry mass. However hard to tell since the values vary greatly between years. If I had to make a guess I would say that mean dry mass has increased over time. This could be changed to by month to look at seasonal variation 


Creating the species richness variable to use for later analysis. 
- not sure now to make this tidy data in my original csv. file so making it a new one.
```{r}
SpeciesRichness <- KelpForest |> 
  group_by(YEAR) |>
   filter(DRY_GM2 > 0)
                    
```
This code made a new data table that only contains the rows of data that indicate a species was present at the time. Thus can use for species richness analysis


Doing some summary statistics on it
```{r}
ggplot(SpeciesRichness, aes(SCIENTIFIC_NAME, YEAR))+
  geom_point()
```
Summary statistics because this graph is impossible to look at. 
```{r}
SpeciesRichness |> group_by(YEAR) |>
  
```


Saving my modified .csv file into the outputs folder for future use. 
```{r}
write.csv(KelpForest,here("Outputs","KelpForestFixed.csv"), row.names=FALSE)
write.csv(SpeciesRichness,here("Outputs","SpechiesRichness.csv"), row.names=FALSE)
```































