---
title: "TotalBiomassAnalysis"
format: html
editor: visual
---

# set up

```{r}
rm(list=ls())
library(tidyverse)
library(here)
library(multcomp)
library(ggfortify)
library(vegan)
library(forecast) #for moving averages
library(gridExtra) #to arrange plots
library(zoo) # for moving averages
library(here)
```

# Reading in data

```{r}
KelpForest <- read.csv(here("Outputs", "KelpForestFixed.csv"))
```

Data has already been checked for proper entry in the data exploration assignment and is fixed in the KelpForestFixed.csv file that I loaded.

# Moving data into long format

using pivot_wider

```{r}
Kelp_Drymass_wide <- KelpForest |> 
  pivot_wider(
    names_from = SCIENTIFIC_NAME, values_from = DRY_GM2) # Using dry mass since I feel it is more representative of actual amount of species than wet mass.
# using scientific name since there are NA values for some of the common names 
  
```

Making the smaller kelp data frame

```{r}
small_kelp <- KelpForest %>% dplyr::select(YEAR:TRANSECT, DENSITY, DRY_GM2, SCIENTIFIC_NAME)
```

Grouping dry mass by species and year

```{r}
kelp_summary <- small_kelp %>% group_by(YEAR, SITE, SCIENTIFIC_NAME) %>% summarise(
  total_dry_mass = sum(DRY_GM2)
) 
```

putting the data in long format

```{r}
Kelp_Summary_Wide <- kelp_summary |> 
  pivot_wider(
    names_from = SCIENTIFIC_NAME, values_from = total_dry_mass
  )
```

I want to do an analysis comparing the total drymass of urchins vs kelp in the forest to see if there are any patterns over time. To do this I am using time series analysis since how their populations compare at specific times is important so a linear regression would not be applicable. 

MY NULL hypothesis is that there will not be a correlation between the kelp and urchin dry mass. 

Since there are multiple species and the names are different i found this grepl() function that will select all rows that contain the words urchin and then kelp. 

# Making the Urchin and Kelp dataframes. 

```{r}
urchins <- KelpForest[grepl(
  "Urchin", KelpForest$COMMON_NAME, 
  ignore.case = TRUE),]
```

Also am filtering for the kingdom of Chromista since there are non kepl species inlcuded that have kelp in their names.  
```{r}
kelp <- KelpForest[grepl(
  "Kelp", KelpForest$COMMON_NAME, 
  ignore.case = TRUE),]
kelp <- kelp |> filter(TAXON_KINGDOM == "Chromista")

```

# Creating time series variables

```{r}
kelp_drymass <- kelp$DRY_GM2

kelp_drymass <- ts(kelp_drymass, frequency = 12, start = c(2008, 21))
```

```{r}
urchin_drymass <- urchins$DRY_GM2

urchin_drymass <- ts(urchin_drymass, frequency = 12, start = c(2008, 21))
```

# Calculating a simple moving average

```{r}
kelp_drymass <- rollmean(kelp_drymass, k=29, fill = NA)
```

```{r}
urchin_drymass <- rollmean(urchin_drymass, k=29, fill = NA)
```

# Putting the time series objects into data frames to be graphed. 

```{r}
kelp$drymass <- as.numeric(kelp_drymass)
```

```{r}
urchins$drymass <- as.numeric(urchin_drymass)
```

# Plotting the data

```{r}
ggplot(kelp, aes(x =DATE))+
  geom_line(aes(y = drymass), color = "blue", linewidth = 1) + #original time series
  geom_line(aes(y = drymass), color = "red", linewidth = 1.2) + #moving average
  labs(title = "KU: k = 29 days",x = "Date", y = "Drymass")+
  theme_bw()
```

```{r}
ggplot(urchins, aes(x =DATE))+
  geom_line(aes(y = drymass), color = "blue", linewidth = 1) + #original time series
  geom_line(aes(y = drymass), color = "red", linewidth = 1.2) + #moving average
  labs(title = "KU: k = 29 days",x = "Date", y = "Drymass")+
  theme_bw()
```

At first glance it is hard to tell if there is any correlation but I think I would like to meet to talk about it there is a way to show that in an actual value. 






























